A Horrocks-type theorem for even orthogonal $K_2$

### 2019-09-28 dl-paper.tex
Gave different proof to proposition 5.18

\documentclass[oneside, 8pt]{amsart}
\usepackage{amscd, amsmath, amssymb, amsthm, amsfonts, amstext, verbatim, mathtools, xfrac, microtype, nameref, thmtools}
\usepackage[breaklinks=false,unicode]{hyperref}
\usepackage[backend=biber, bibencoding=utf8, giveninits=true, citestyle=numeric-comp, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}
\usepackage[capitalize]{cleveref}
\usepackage[matrix,arrow,curve]{xy}
\usepackage{tikz}
\usepackage{enumitem}
%\usepackage[notref,notcite]{showkeys}

@@

For a subset of roots $U \subseteq \Phi$ we denote by $\mathcal{Z}(U, R, I)$ the subset of roots consisting of elements $x_\alpha(s)$, $s \in I$, $\alpha \in \Phi$ and $z_\alpha(s, \xi)$, $\alpha \in U$, $s\in I$, $\xi \in R$.

@@

Recall that a closed root subset $S \subseteq \Phi$ is called {\it parabolic} (resp. {\it reductive}, resp. {\it special}) if $S \cup -S = \Phi$ (resp. $S = -S$, resp. $S \cap (-S) = \varnothing$).
The {\it special part} $\Sigma_S$ of a parabolic subset $S$, by definition, consists of all $\alpha \in S$ such that $-\alpha \not\in S$.
\begin{externaltheorem}[Stepanov] \label{thm:Stepanov} 
Let $S \subseteq \Phi$ be a parabolic subset of $\Phi$. Then the group $\overline{\St}(\Phi, R, I)$ is generated by the set $\mathcal{Z}(\Sigma_S, R, I)$.
 \end{externaltheorem} \begin{proof} See~\cite[Lemma~4]{S15}. \end{proof}

@@

\begin{lemma} \label{Crels}
The elements $c_\alpha(s, t)$ satisfy the following relations for all $s\in I,\ t\in J,\ \xi\in R$.
 \begin{enumerate}
 \item \label{C1} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = x_{\alpha}(- st\xi) \cdot x_{\alpha+\beta}(N_{\alpha,\beta}\cdot s^2t\xi)$ if $\alpha+\beta \in \Phi$;
 \item \label{C2} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = x_{\alpha}(st\xi + s^2t^2\xi) \cdot x_{\alpha-\beta}(N_{-\alpha, \beta}\cdot st^2\xi)$ if $\alpha-\beta \in \Phi$;  
 \item \label{C3} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = 1$ if $\alpha \perp \beta$;  
 \item \label{C4} If $\alpha+\beta\in\Phi$ then holds:
  \[c_{\alpha+\beta}(s, t\xi) = [x_{\beta}(st),\ x_{-\beta}(\xi)] ^ {x_{\alpha+\beta}(-s) x_{-\alpha}(\epsilon t)} \cdot c_{\alpha}(\epsilon s\xi, -\epsilon t)^{-1} \cdot x_{-\beta}(-st\xi^2),\]
  where $\epsilon = N_{\alpha,\beta}$.
 \end{enumerate}
\end{lemma}
\begin{proof}
The first assertion follows from the following computation:
\begin{multline} \nonumber [c_{\beta}(s, t),\ x_{\alpha}(\xi)]  = [x_{\beta}(s),\ x_{-\beta}(t)] \cdot [x_{-\beta}(t),\ x_{\beta}(s)\cdot x_{\alpha + \beta}(N_{\alpha, \beta} \cdot s\xi)] = \\
 = {}^{x_\beta(s)}\![x_{-\beta}(t),\ x_{\alpha+\beta}(N_{\alpha, \beta} \cdot s\xi)] = x_{\alpha}(- st\xi) \cdot x_{\alpha+\beta}(N_{\alpha,\beta} \cdot s^2t\xi). \end{multline}

@@

\begin{corollary} \label{Kgen-strong} For a local pair $(A, M)$ and arbitrary fixed root $\gamma$ of an irreducible simply-laced root system $\Phi$ the subgroup $K(A[X], M[X])$ is generated as a group by $\overline{\St}(\Phi, A[X], XM[X])$ and the elements $c_{\gamma}(m, X\eta)$, where $m \in M$, $\eta \in A[X]$. \end{corollary}
\begin{proof} Substituting $\xi = 1$, $s = m$, $t = X\eta$ into relation~\eqref{C4} of~\cref{Crels} we obtain that modulo 
 $\overline{\St}(\Phi, A[X], XM[X])$ the element $c_{\alpha + \beta}(m, X\eta)$ is equivalent to $c_{\alpha}(-\epsilon m, -\epsilon X \eta)^{-1}$, $\epsilon = N_{\alpha, \beta}$. The assertion of the corollary now easily follows from the irreducibility of $\Phi$. \end{proof}

@@

\subsection{The subgroups $K(\alpha, \beta)$ and their properties} \label{sec:Pa0-Steinberg}
Throughout this subsection $\alpha, \beta$ denote a fixed pair of roots of $\Phi$ forming a sharp angle (i.\,e. $\langle \alpha, \beta \rangle = 1$). We denote by $\Psi$ the subsystem of type $\rA_2$ generated by $\alpha$ and $\beta$.
\begin{df} Denote by $K(\alpha, \beta)$ the subgroup of $\St(\Phi, A[X, X^{-1}])$ generated by the following elements (as before, $\xi \in A[X]$, $f\in M[X]$):
 \begin{enumerate}
  \item $z_\gamma(Xf, \xi)$, for all $\gamma \in \Phi \setminus \Psi$;
  \item $x_{-\alpha}(X^2f)$, $x_{-\beta}(X^2f)$;
  \item $x_{\alpha}(Xf)$, $x_\beta(Xf)$;
  \item $x_{\alpha-\beta}(Xf)$, $x_{\beta-\alpha}(Xf)$;
  \item $c_{\delta}(f, X\xi)$ for some fixed root $\delta \in \Phi \setminus \Psi$. \end{enumerate} \end{df}

\begin{prop} \label{K-a-b} Every element $g \in K(A[X], M[X])$ can be written as a product $g_0 \cdot x_{-\alpha}(mX) \cdot x_{-\beta}(m'X)$ for some $m, m' \in M$ and $g_0 \in K(\alpha, \beta)$. 
Consequently, $K(\alpha, \beta)$ coincides with the intersection $K(A[X], M[X]) \cap P_\alpha(0) \cap P_\beta(0)$. 
In particular, $K(\alpha, \beta)$ does not depend on the choice of the root $\delta$. \end{prop}
\begin{proof} Denote by $\Sigma$ the special part of the parabolic subset $S = \Psi \cup \Phi^+$. 
Applying Stepanov theorem to $S$ we obtain that every element of $\St(\Phi, A[X], XM[X])$ can be presented as a product of $z_\gamma(Xf, \xi)$, 
$\gamma \in \Sigma(S) \subseteq \Phi\setminus \Psi$ and $x_\gamma(Xf)$, $\gamma \in \Psi$. By~\cref{Kgen-strong} we can write $g$ as a product of these generators and generators of type 5.

 The next step of the proof is to decompose each factor of the form $x_{\epsilon}(Xf)$, $\epsilon \in \{-\alpha,-\beta\}$ appearing in this product as
 $x_{\epsilon}(X^2f') \cdot x_{\epsilon}(mX)$ and then move $x_{\epsilon}(mX)$ to the rightmost position conjugating by $x_{\epsilon}(mX)$ all factors along the way.
 Notice that no terms of the form $x_{\epsilon}(mX)$, $\epsilon \in \{ -\alpha, -\beta \}$ may appear after simplification of these conjugates, so this process is guaranteed to terminate.
 To see this notice that the extra terms in the right-hand sides of \eqref{eq3-1} and~\eqref{eq3-3} 
 (as well as all terms in the right-hand sides of~\eqref{C1},\eqref{C2} of~\cref{Crels} if one substitutes corresponding values into them) have terms of degree at least $2$ in $X$.

 As a side effect of this transformation, however, factors $x_\alpha(Xf)$ and $x_\beta(Xf)$ appearing in $g$ might transform into factors of the form $z_\alpha(Xf, X\xi)$, and $z_\beta(Xf, X\xi)$. We can express each factor $z_\alpha(Xf, X\xi)$ through the generators of $K(\alpha, \beta)$ as follows. Pick a root $\gamma \in \Phi\setminus \Psi$ forming an obtuse angle with $\alpha$ and then invoke~\eqref{eq:zalpha}:
 \begin{multline} z_\alpha(Xf, X\xi) = x_{\alpha+\gamma}(\epsilon X^2f) \cdot x_{\gamma}(-X^3 \xi f) \cdot x_{-\gamma}(Xf\xi) \cdot x_\alpha(Xf) \cdot \\ 
 \cdot z_{\alpha+\gamma}(-\epsilon X^2 f, -\epsilon \xi) \cdot z_{-\gamma}(-Xf\xi, -X) \cdot x_{-\alpha-\gamma}(-\epsilon X^2\xi^2 f) \cdot x_{-\alpha}(-X^3 f \xi^2),\ \text{for } \epsilon = N_{\alpha, \gamma}. \end{multline}
 The expression in the right-hand side is a product of generators of type 1, 1, 1, 3, 1, 1, 1, 2. Similar argument also applies to $z_\beta(Xf, X\xi)$.

 After collecting terms $x_\epsilon(mX)$, $\epsilon \in \{ -\alpha, -\beta \}$ in the right-hand side of the resulting expression we obtain the desired decomposition for $g$.

>>>>

\documentclass[oneside, 8pt]{amsart}
\usepackage{amscd, amsmath, amssymb, amsthm, amsfonts, amstext, verbatim, mathtools, xfrac, microtype, nameref, thmtools}
\usepackage[breaklinks=false,unicode]{hyperref}
\usepackage[backend=biber, bibencoding=utf8, giveninits=true, citestyle=numeric-comp, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}
\usepackage[capitalize]{cleveref}
\usepackage[matrix,arrow,curve]{xy}
\usepackage{tikz}
\usepackage{enumitem}
\usepackage[notref,notcite]{showkeys}

@@

@@

Recall that a closed root subset $S \subseteq \Phi$ is called {\it parabolic} (resp. {\it reductive}, resp. {\it special}) if $S \cup -S = \Phi$ (resp. $S = -S$, resp. $S \cap (-S) = \varnothing$).
The {\it special part} $\Sigma_S$ of a parabolic subset $S$, by definition, consists of all $\alpha \in S$ such that $-\alpha \not\in S$.

For a subset of roots $U \subseteq \Phi$ we denote by $\mathcal{Z}(U, R, I)$ the subset of roots consisting of elements $x_\alpha(s)$, $s \in I$, $\alpha \in \Phi$ and $z_\alpha(s, \xi)$, $\alpha \in U$, $s\in I$, $\xi \in R$.
\begin{externaltheorem}[Stepanov] \label{thm:Stepanov} 
Let $S \subseteq \Phi$ be a parabolic subset of $\Phi$. Then the group $\overline{\St}(\Phi, R, I)$ is generated by the set $\mathcal{Z}(\Sigma_S, R, I)$.
 \end{externaltheorem} \begin{proof} See~\cite[Lemma~4]{S15}. \end{proof}

@@

\begin{lemma} \label{Crels}
The elements $c_\alpha(s, t)$ satisfy the following relations for all $s\in I,\ t\in J,\ \xi\in R$.
 \begin{enumerate}
 \item \label{C1} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = x_{\alpha}(- st\xi) \cdot x_{\alpha+\beta}(N_{\alpha,\beta}\cdot s^2t\xi)$ if $\alpha+\beta \in \Phi$;
 \item \label{C2} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = x_{\alpha}(st\xi + s^2t^2\xi) \cdot x_{\alpha-\beta}(N_{-\alpha, \beta}\cdot st^2\xi)$ if $\alpha-\beta \in \Phi$;  
 \item \label{C3} $[c_\beta(s, t),\ x_{\alpha}(\xi)] = 1$ if $\alpha \perp \beta$;  
 \item \label{C4} If $\alpha+\beta\in\Phi$ then holds:
  \begin{equation} \label{eq:C4} c_{\alpha+\beta}(s, t\xi) = [x_{\beta}(st),\ x_{-\beta}(\xi)] ^ {x_{\alpha+\beta}(-s) x_{-\alpha}(\epsilon t)} \cdot c_{\alpha}(\epsilon s\xi, -\epsilon t)^{-1} \cdot x_{-\beta}(-st\xi^2),\end{equation}
  where $\epsilon = N_{\alpha,\beta}$.
 \end{enumerate}
\end{lemma}
\begin{proof}
The first assertion follows from the following computation:
\begin{multline} \nonumber [c_{\beta}(s, t),\ x_{\alpha}(\xi)]  = [x_{\beta}(s),\ x_{-\beta}(t)] \cdot [x_{-\beta}(t),\ x_{\beta}(s)\cdot x_{\alpha + \beta}(N_{\alpha, \beta} \cdot s\xi)] = \\
 = {}^{x_\beta(s)}\![x_{-\beta}(t),\ x_{\alpha+\beta}(N_{\alpha, \beta} \cdot s\xi)] = x_{\alpha}(- st\xi) \cdot x_{\alpha+\beta}(N_{\alpha,\beta} \cdot s^2t\xi). \end{multline}

@@

\begin{corollary} \label{Kgen-strong} For a local pair $(A, M)$ and arbitrary fixed root $\gamma$ of an irreducible simply-laced root system $\Phi$ the subgroup $K(A[X], M[X])$ is generated as a group by $\overline{\St}(\Phi, A[X], XM[X])$ and the elements $c_{\gamma}(m, X\eta)$, where $m \in M$, $\eta \in A[X]$. \end{corollary}
\begin{proof} Substituting $\xi = 1$, $s = m$, $t = X\eta$ into~\eqref{eq:C4} we obtain that modulo 
 $\overline{\St}(\Phi, A[X], XM[X])$ the element $c_{\alpha + \beta}(m, X\eta)$ is equivalent to $c_{\alpha}(-\epsilon m, -\epsilon X \eta)^{-1}$, $\epsilon = N_{\alpha, \beta}$. The assertion of the corollary now easily follows from the irreducibility of $\Phi$. \end{proof}

@@

\subsection{The subgroups $K(\alpha, \beta)$ and their properties} \label{sec:Pa0-Steinberg}
Throughout this subsection $\alpha, \beta$ denote a fixed pair of roots of $\Phi$ forming a sharp angle (i.\,e. $\langle \alpha, \beta \rangle = 1$). We denote by $\Psi$ the subsystem of type $\rA_2$ generated by $\alpha$ and $\beta$.
\begin{df}\label{def:Kab} Denote by $\widetilde{K}(\alpha, \beta)$ the subgroup of $\St(\Phi, A[X], M[X])$ generated by the following five families of elements (below $\xi \in A[X]$, $f\in M[X]$):
 \begin{enumerate}
  \item $z_\gamma(Xf, \xi)$, for all $\gamma \in \Phi \setminus \Psi$;
  \item $x_{-\alpha}(X^2f)$, $x_{-\beta}(X^2f)$;
  \item $x_{\alpha}(Xf)$, $x_\beta(Xf)$;
  \item $x_{\alpha-\beta}(Xf)$, $x_{\beta-\alpha}(Xf)$;
  \item $c_{\delta}(f, X\xi)$ for some fixed root $\delta \in \Phi \setminus \Psi$. \end{enumerate}
Denote by $K(\alpha, \beta)$ the image of $\widetilde{K}(\alpha, \beta)$ in $\St(\Phi, A[X, X\inv])$. \end{df}

\begin{prop} \label{K-a-b} Every element $g \in K(A[X], M[X])$ can be written as a product $g_0 \cdot x_{-\alpha}(mX) \cdot x_{-\beta}(m'X)$ for some $m, m' \in M$ and $g_0 \in \widetilde{K}(\alpha, \beta)$. 
Consequently, $K(\alpha, \beta)$ coincides with the intersection $\mathrm{Im}(K(A[X], M[X]) \to \St(\Phi, A[X, X\inv])) \cap P_\alpha(0) \cap P_\beta(0)$. In particular, $K(\alpha, \beta)$ does not depend on the choice of the root $\delta$. \end{prop}
\begin{proof} 
For $i=1..5$ denote by $\mathcal{G}_i$ the set of all generators of type $i$ from~\cref{def:Kab}.
By~\cref{thm:Stepanov} every element of $\St(\Phi, A[X], XM[X])$ can be presented as a product of generators $\mathcal{G}_1$ and generators $\mathcal{G}' = \{ x_{\gamma}(Xf) \mid \gamma \in \Psi\}$. In turn, by~\cref{Kgen-strong} one can express $g$ as a product of these generators and generators $\mathcal{G}_5$. %We are going to rewrite this product so that the generators of $\widetilde{K}(\alpha, \beta)$ accumulate on the left-hand side of this expression, while the elements of the root subgroups $X_{-\alpha}(MX)$ and $X_{-\beta}(MX)$ accumulate on the right-hand side.

By~\cref{thm:Stepanov} $\widetilde{K}(\alpha, \beta)$ contains the normal subgroup $G_0 = \St(\Phi, A[X], X^2M[X])$, so it suffices to obtain the required presentation of $g$ modulo $G_0$. By~\cref{Zrels} and~\cref{Crels} the elements of $\mathcal{G}'$ commute with the elements of $\mathcal{G}_1$ and $\mathcal{G}_5$ modulo $G_0$, therefore we can rewrite $g$ as $g_1 \cdot g_2$, where $g_1 \in \langle \mathcal{G}_1, \mathcal{G}_5, G_0 \rangle \subseteq \widetilde{K}(\alpha, \beta)$ and $g_2 \in \langle \mathcal{G}' \rangle$.

Now let us study the structure of the subgroup $\langle \mathcal{G}' \rangle$. It is clear that  $x_{\gamma_1}(mX)$ and $x_{\gamma_2}(m'X)$ commute modulo $G_0$ whenever $\gamma_1 \neq - \gamma_2$, $\gamma_1, \gamma_2 \in \Psi$. On the other hand, for $\gamma \in \Psi$ any commutator $c_{\gamma}(mX, m'X)$ is congruent to some generator of $\mathcal{G}_5$  modulo $G_0$ (cf. the proof of~\cref{Kgen-strong}) and, therefore, commutes with the elements of $\mathcal{G}'$ modulo $G_0$.

Now pick each factor $x_{\gamma}(Xf)$, $\gamma \in \{ -\alpha, -\beta \}$ appearing in the presentation for $g_2$, decompose it as $x_{\gamma}(X^2f') \cdot x_{\gamma}(m'X)$ for some $f', m'$ and then move the second factor to the rightmost position within $g_2$ conjugating all factors along the way. By the previous paragraph only elements of  $\mathcal{G}_5$ may appear modulo $G_0$ after the simplification of these conjugates.
Thus, we can rewrite $g_2$ as $g_{21} \cdot g_{22}$, where $g_{21} \in \langle \mathcal{G}_2, \mathcal{G}_3, \mathcal{G}_4, \mathcal{G}_5, G_0 \rangle \subseteq \widetilde{K}(\alpha, \beta)$ and $g_{22} \in \langle X_{-\alpha}(MX), X_{-\beta}(MX) \rangle = X_{-\alpha}(MX) \cdot X_{-\beta}(MX)$.
Thus, we have obtained the desired decomposition for $g$.
